Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant [098051].}

### 2016-04-20 paper/paper.tex
Proofreading and pruning the paper; some changes to benchmarks.

\begin{abstract}
Variation graphs, which represent genetic variation within a population, are replacing sequences as reference genomes. Path indexes are one of the most important tools for working with variation graphs. They generalize text indexes for graphs, allowing one to find the paths matching the query string. We propose using pruned de Bruijn graphs as path indexes, and encode them space-efficiently with the Burrows-Wheeler transform. We also generalize many ideas from text indexing literature to work with graphs. The proposed approach has been implemented in the variation graph toolkit vg.
\end{abstract}

@@

Sequencing pipelines typically start with mapping the reads from the sequenced genome to a \emph{reference genome} of the same species. As reference genomes are usually assembled from the genomes of a small number of individuals, they are biased towards those individuals. This \emph{reference bias} may affect the results of subsequent analysis, especially when the sequenced individuals are from different populations than the ones behind the reference genome.

Recently there has been much interest in \emph{variation graphs} (also called pan-genomes, graph genomes, or reference graphs), which encode the genetic variation within a population as a graph \cite{Paten2014,Marcus2014,Church2015,Dilthey2015,Marschall2016}. These graphs are eventually expected to replace reference sequences as reference genomes. The shift to graph references will likely take years, as new methods and tools must be developed to replace those based on linear references.

The \emph{variation graph toolkit vg} \cite{Garrison2014-2016} is a community effort to develop such tools. This paper describes \emph{GCSA2}, the path index developed for vg. A \emph{path index} is a generalization of text indexes for labeled graphs. They are used for finding paths with labels that match the query string. Indexing graphs for path queries is an inherently hard problem, as the number of paths increases exponentially with path length. Hence the design of a path index is a trade-off between maximum supported query length, the size of the index, and pruning complex regions of the graph before indexing.

Mapping reads to a graph was first investigated by Schneeberger et~al.~\cite{Schneeberger2009}. Sirén et~al.~\cite{Siren2014} developed \emph{GCSA}, which generalized the FM\nobreakdash-index \cite{Ferragina2005a} from a text index to a path index. The original GCSA could only index directed acyclic graphs and depended on pruning the complex regions, as it did not place any limits on query length. Kim et~al.~\cite{Kim2015-2016} combined GCSA with the ideas developed for the HISAT read aligner \cite{Kim2015}. The resulting HISAT2 aligner was the first practical graph-based read aligner, though it only uses the graph for more accurate mapping to a linear reference.

The \emph{succinct de~Bruijn graphs} of Bowe et.~al.~\cite{Bowe2012} used a similar generalization of the FM\nobreakdash-index as in GCSA for representing \emph{de~Bruijn graphs} space-efficiently. Rødland \cite{Roedland2013} proposed another similar generalization. The succinct de~Bruijn graphs have been generalized to represent \orderk{k} de~Bruijn graphs with different values of $k$ in the same structure \cite{Boucher2014}, but they still need to store the graph for the largest value of $k$ explicitly. Compacted \cite{Cazaux2014} and compressed de~Bruijn graphs \cite{Marcus2014} achieve similar space savings by representing unary paths in the graph as a single node. A prototype read aligner based on compacted de~Bruijn graphs has been proposed \cite{Limasset2015}.

Other approaches to indexing variation graphs include BWBBLE \cite{Huang2013}, which indexes the graph as a collection of sequences. Simple variants are encoded by using the powerset alphabet, while more complex variants are represented as separate sequences, with a sufficient amount of context around the variant site. This is usually enough to keep the index small, at the price of much higher query times. The hypertext index \cite{Thachuk2013} is another similar idea. The nodes of the graph are labeled with strings, which are indexed using an ordinary FM-index. Partial matches in the strings are combined into full matches with range queries in the edge matrix. While matches crossing one edge are easy to find, the approach becomes complex and potentially slow with matches crossing multiple edges. There are also several indexes (e.g.~\cite{Huang2010,Wandelt2013,Danek2014,Na2015}), which use a graph as a space-efficient representation for similar sequences but index the sequences instead of the graph.

@@

A \emph{string} $S[0, n-1] = s_{0} \dotsm s_{n-1}$ of length $\abs{S} = n$ is a sequence of \emph{characters} over an \emph{alphabet} $\Sigma = \set{0, \dotsc, \sigma - 1}$. For indexing purposes, we often consider \emph{text} strings $T[0, n-1]$ terminated by an \emph{endmarker} $T[n-1] = \$ = 0$ not found anywhere else in the text. \emph{Binary} sequences are strings over the alphabet $\set{0, 1}$. A \emph{substring} of string $S$ is a sequence of the form $S[i, j] = s_{i} \dotsm s_{j}$. We call substrings of the type $S[0, j]$ and $S[i, n-1]$ \emph{prefixes} and \emph{suffixes}, respectively, and substrings of length $k$ as \kmer{k}s. Substring $S[i, j]$ is a \emph{proper} substring of string $S$, if $S \ne S[i, j]$. We say that string $S'$ is a substring of string collection $\mathcal{S}$, if there is a string $S \in \mathcal{S}$ such that $S'$ is a substring of $S$.

@@

Given string $S[0, n-1]$, we define $S.\rank(i, c)$ to be the number of occurrences of character $c$ in the prefix $S[0, i-1]$. We also define $S.\select(i, c) = \arg \max_{j \le n} S.\rank(j, c) < i$ as the position of the occurrence of character $c$ with rank $i > 0$.\footnote{These definitions are used in the SDSL library \cite{Gog2014b}. We assume for convenience that $S.\select(0, c) = -1$.} A \emph{bitvector} is a binary sequence supporting efficient $\rank$/$\select$ queries. \emph{Wavelet trees} \cite{Grossi2003} are space-efficient data structures that use bitvectors to support $\rank$/$\select$ queries on arbitrary strings.

@@

The \emph{suffix tree} \cite{Weiner1973} is the most fundamental full-text index supporting substring queries. It is formed by taking the suffixes of the text, storing them in a trie, and compacting unary paths in the trie into single edges. If $v$ is a node of a suffix tree, we write $\ell(v)$ to denote the label of the path from the root to node $v$. Although fast and versatile, suffix trees are only of limited use in indexing large texts, as they require much more space than the text itself.

\emph{Suffix arrays} \cite{Manber1993} were introduced as a space-efficient alternative to the suffix tree. The suffix array of text $T[0, n-1]$ is an array of pointers $\SA[0, n-1]$ to the suffixes of the text in \emph{lexicographic order}. Given a text and its suffix array, we can find the occurrences of \emph{pattern} string $X$ in the text by using \emph{binary search} in $\Oh{\abs{X} \log n}$ time. The suffix array requires $n \log n$ bits of space in addition to the $n \log \sigma$ bits used by the text --- still much more than the text --- while its functionality is more limited than that of the suffix tree. See Figure~\ref{figure:indexes} for an example of the suffix array and related structures.

@@

The \emph{Burrows-Wheeler transform (BWT)} \cite{Burrows1994} is a permutation of the text with the same combinatorial structure as the suffix array. Given text $T[0, n-1]$, its Burrows-Wheeler transform is a string $\BWT[0, n-1]$, where $\BWT[i] = T[(\SA[i]-1) \bmod n]$. Given the \emph{lexicographic rank} $i$ of suffix $T[\SA[i], n-1]$, we can use \emph{\LFmapping} on the BWT to find the lexicographic rank of the previous suffix $T[(\SA[i]-1) \bmod n, n-1]$. Let
$$
\LF(i) = \Carray[\BWT[i]] + \BWT.\rank(i, \BWT[i]),
$$
where $\Carray[c]$ is the number of occurrences of characters $c' < c$ in the BWT. Then $\SA[\LF(i)] = (\SA[i]-1) \bmod n$. We can also generalize the definition for any character $c \in \Sigma$:
$$
\LF(i, c) = \Carray[c] + \BWT.\rank(i, c).
$$
Let $X$ be a string. If there are $i$ suffixes $S'$ of text $T$ such that $S' < X$ in lexicographic order, then there are $\LF(i, c)$ suffixes smaller than string $cX$ in lexicographic order.

Given the similarities to the suffix array, we can use the BWT as a space-efficient text index. The \emph{\FMindex} \cite{Ferragina2005a} combines a plain or compressed representation of the BWT supporting $\rank$/$\select$ queries, the $\Carray$ array, and a set of \emph{sampled pointers} from the suffix array. It finds the \emph{lexicographic range} of suffixes matching pattern $X$ (having $X$ as a prefix) with a process called \emph{backward searching}. If the lexicographic range matching suffix $X[i+1, \abs{X}-1]$ of the pattern is $\SA[sp, ep]$, then the lexicographic range matching suffix $X[i, \abs{X}-1]$ of the pattern is $\SA[\LF(sp, X[i]), \LF(ep+1, X[i]) - 1]$. Finding the lexicographic range matching the entire pattern requires $\Oh{\abs{X}}$ rank queries.

We can use the sampled suffix array pointers to find the text positions containing the occurrences. If $\SA[i]$ is not sampled, we start iterating $\LF(i)$, until we find a sampled pointer. If we find a sample at $\SA[\LF^{k}(i)]$, we know that
$$
\SA[i] = (\SA[\LF^{k}(i)] + k) \bmod n.
$$
If we have sampled one out of $d$ suffix array pointers at regular intervals, finding each occurrence takes $\Oh{d}$ rank queries. If we also sample one out of $d'$ \emph{inverse suffix array} pointers\footnote{The suffix array is a permutation of $\set{0, \dotsc, n-1}$, and the inverse suffix array is the inverse permutation.}, we can \emph{extract} an arbitrary substring $X$ of the text using $\Oh{\abs{X}+d'}$ rank queries.

@@

The graphs we consider are \emph{labeled} with alphabet $\Sigma$: each node $v \in V$ has a \emph{label} $G.\glabel(v) \in \Sigma$. A \emph{path} in a graph is a sequence of nodes $P = v_{0} \dotsm v_{\abs{P}-1}$ such that $(v_{i}, v_{i+1}) \in E$ for all $i, i+1 \in \set{0, \dotsc, \abs{P}-1}$. We say that $v_{0}$ is the \emph{start} node and $v_{\abs{P}-1}$ is the \emph{end} node of the path. The label of a path is the concatenation of node labels $G.\glabel(P) = G.\glabel(v_{0}) \dotsm G.\glabel(v_{\abs{P}-1})$.

@@

We assume for convenience that all graphs have two special nodes: the \emph{source} node $s$ and the \emph{sink} node $t$. To distinguish them from the other nodes, we label them with characters $G.\glabel(s) = \#$ and $G.\glabel(t) = \$$. The label of the sink node is unique in the graph, while the label of the source node can be used in other nodes for technical purposes. We add an edge $(s, v)$ for all nodes $v \in V \setminus \set{s}$ that have no other incoming edges, and an edge $(v, t)$ for all nodes $v \in V \setminus \set{t}$ with no other outgoing edges. We also add edge $(t, s)$ to guarantee that $G.\gindegree(v) \ge 1$ and $G.\goutdegree(v) \ge 1$ for all nodes $v \in V$. However, this edge is not considered a real edge, and no path can cross it.

@@

In this paper, we work with de~Bruijn graphs and their generalizations. In order to do so, we need to define collections of (finite or infinite) sequences suitable for constructing \orderk{k} de~Bruijn graphs.

@@

De~Bruijn graphs have several properties that make them useful for indexing purposes. Node keys are prefixes of the labels of all paths of length at least $k$ starting from the node. This makes it possible to sort the nodes unambiguously by path labels. Every substring of the \kcollection{k} is the label of a path in the de~Bruijn graph, and every path label of length at most $k+1$ is a substring of the collection. In Section~\ref{sect:path-indexes}, we develop an index structure based on a generalization of de~Bruijn graphs.

@@

The \emph{generalized compressed suffix array} (GCSA) \cite{Siren2014} is a further generalization of the XBW transform for a class of graphs that includes \emph{directed acyclic graphs} and de Bruijn graphs. Before a graph can be indexed, we have to transform it into an equivalent graph, where the nodes can be unambiguously sorted by the labels of the right-maximal paths starting from them. In the worst case, the transformation increases the size of the graph exponentially. After sorting the nodes of the transformed graph in lexicographic order, we encode them using three sequences. $\BWT$ contains the labels of the predecessors, while bitvectors $\bvIN$ and $\bvOUT$ encode the indegrees and outdegrees of each node in the same way as bitvector $B$ of the XBW transform. \LFmapping{} uses $\select$ queries on bitvector $\bvIN$ to map nodes to BWT ranges, ordinary \LFmapping{} with $\BWT$ to map incoming edges to the corresponding outgoing edges, and $\rank$ queries on bitvector $\bvOUT$ to map the outgoing edges to the predecessor nodes.

@@

The design of a path index is a trade-off between the maximum supported query length and the size of the index. In the worst case, a graph may have up to $\sigma^{k}$ distinct labels for paths of length $k$ (ignoring the special treatment of characters $\#$ and $\$$), and the length of possible paths is unlimited in cyclic graphs. GCSA \cite{Siren2014} tried to avoid this by considering only a relatively simple class of graphs, using \emph{pruning heuristics} to simplify the graphs when necessary. The price of this was the possibility of \emph{false negatives}: path labels that exist in the graph but not in the index.

@@

Another representation of the \kmer{k} index trades query performance for the ability to search for shorter patterns. Instead of using a hash table, we sort the key-value pairs by their keys, and store the pairs in array $A$ in sorted order. Given a pattern $X \in \Sigma^{\ast}$, $\abs{X} \le k$, we use binary search to find the range $A[sp, ep]$ of pairs $(X_{i}, V_{i})$, where pattern $X$ is a prefix of the key $X_{i}$. We then list the union $\bigcup_{i=sp}^{ep} V_{i}$ as the set of matches.

As de~Bruijn graphs and \kmer{k} indexes are both based on \kmer{k}s, we can use one to simulate the other. In order to define the de~Bruijn graph of a graph, we build a \kcollection{k} based on the collection $\mathcal{S}$ of the labels of the maximal paths in graph $G = (V, E)$. If sequence $S \in \mathcal{S}$ is the label of path $P = (v_{i})_{i \in Z}$ in the graph, we set $\mathcal{S}.\gnode(S, i) = v_{i}$ for all positions $i \in Z$. We then transform $\mathcal{S}$ into a \kcollection{k} by inserting the required the number of characters $\#$ to the beginning of each non-left-infinite sequence, and characters $\$$ to the end of each non-right-infinite sequence. If $S[i]$ is a $\#$ we inserted, we set $\mathcal{S}.\gnode(S, i) = s:j$, where $s$ is the start node of $G$ and $j$ is the distance to the nearest non-inserted $\#$ in $S$. Similarly, if $S[i]$ is an inserted $\$$, we set $\mathcal{S}.\gnode(S, i) = (t, j)$, where $t$ is the sink node and $j$ is the distance to the nearest non-inserted $\$$.

\begin{definition}[de~Bruijn graph of a graph]
Let $G$ be a labeled graph, and let $\mathcal{S}$ be the \kcollection{k} of maximal path labels in $G$. The \orderk{k} de~Bruijn graph of $\mathcal{S}$ is the \orderk{k} de~Bruijn graph of graph $G$.
\end{definition}

See Figure~\ref{figure:graph-dbg} for an example of a de~Bruijn graph of a graph. We can use a \kmer{k+1} index of graph $G = (V, E)$ to simulate the \orderk{k} de~Bruijn graph $G' = (V', E')$ of $G$. As the nodes of the de~Bruijn graph correspond to \kmer{k}s and the edges correspond to \kmer{k+1}s, we can represent the nodes by the adjacent edges. If we are interested in node $v \in V'$ with $G'.\gkey(v) = X$, we search for incoming edges $cX$ and outgoing edges $Xc$ for all $c \in \Sigma$. Determining the existence of a node requires $2 \sigma$ queries in the \kmer{k+1} index, and we learn the presence of adjacent nodes and edges in the process.

@@

On the other hand, we can use the \orderk{k} de~Bruijn graph $G' = (V', E')$ of graph $G = (V, E)$ as a \kmer{k} index of $G$ that supports queries of arbitrary length. Let $\mathcal{S}$ be the \kcollection{k} used for building the de~Bruijn graph. For each node $v \in V'$, we attach a set of nodes of graph $G$ as a \emph{value}: $G'.\gvalue(v) = \set{ \mathcal{S}.\gnode(S, i) \mid S \in \mathcal{S}, S[i, i+k-1] = G'.\gkey(v)}.$ Apart from some technicalities near the source/sink nodes, set $G'.\gvalue(v)$ is the set of start nodes of the paths $P$ in graph $G$ with $G.\glabel(P) = G'.\gkey(v)$. Due to the nature of de~Bruijn graphs, this index will not produce any false negatives. There may be \emph{false positives} (path labels that exist in the index but not in the input graph) with patterns longer than $k$, but we can avoid them by \emph{verifying} the results of such queries in the input graph.

@@

De~Bruijn graphs are a special case of path graphs. They represent the paths of length $k$ in the \emph{input graph} as nodes and the pairs of paths overlapping on $k-1$ nodes of the input graph as edges. We can generalize the notion by allowing the nodes of the path graph to represent paths of different lengths.

\begin{definition}[Path graph]
Let $\mathcal{S}$ be a \kcollection{k} of the labels of maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of strings of length at most $k$ containing $\#^{k}$ and $\$^{k}$. We assume that each right-maximal substring $S$ in $\mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$, and that each string $K \in \mathcal{K}$ is a substring of $\mathcal{S}$. The \orderk{k} \emph{path graph} of graph $G$ with \emph{key set} $\mathcal{K}$ is a graph $G' = (V', E')$ such that
\begin{itemize}
\item each node $v_{K} \in V'$ represent a distinct key $K \in \mathcal{K}$, with $G'.\glabel(v_{K}) = K[0]$;
\item each node has a key $G'.\gkey(v_{K}) = K$ and a value $G'.\gvalue(v_{K}) = V_{K}$, where $V_{K}$ is the set of nodes $\mathcal{S}.\gnode(S, i) \in V$ for $S \in \mathcal{S}$ and positions $i$ such that $S[i, i+\abs{K}-1] = K$; and
\item each edge $(v_{K}, v_{K'}) \in E'$ represents the occurrence of substring $K[0] K'$ in $\mathcal{S}$ such that strings $K$ and $K[0] K'$ prefix-match.
\end{itemize}
We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
\end{definition}

>>>>

\begin{abstract}
Variation graphs, which represent genetic variation within a population, are replacing sequences as reference genomes. Path indexes are one of the most important tools for working with variation graphs. They generalize text indexes for graphs, allowing one to find the paths matching the query string. We propose using pruned de Bruijn graphs as path indexes, and encode them space-efficiently with the Burrows-Wheeler transform. We also generalize ideas from text indexing literature and use them with graphs. The proposed approach has been implemented in the variation graph toolkit vg.
\end{abstract}

@@

Sequence analysis pipelines typically start with mapping the reads from the sequenced genome to a \emph{reference genome} of the same species. As reference genomes are usually assembled from the genomes of a small number of individuals, they are biased towards those individuals. This \emph{reference bias} may affect the results of subsequent analysis, especially when the sequenced individuals are from different populations than the ones behind the reference genome.

\emph{Variation graphs} (also called pan-genomes, graph genomes, or reference graphs), which encode the genetic variation within a population as a graph, have been proposed as a solution to the reference bias \cite{Paten2014,Marcus2014,Church2015,Dilthey2015,Marschall2016}. These graphs are eventually expected to replace reference sequences as reference genomes. The shift to graph references will likely take years, as new methods and tools must be developed to replace those based on linear references.

The \emph{variation graph toolkit vg} \cite{Garrison2014-2016} is a community effort to develop such tools. This paper describes \emph{GCSA2}, the path index used in vg. A \emph{path index} is a generalization of text indexes for labeled graphs. Given a query string, they find the paths with a label matching the query. Indexing graphs for path queries is inherently hard, as the number of paths increases exponentially with path length. The design of a path index is hence a trade-off between maximum query length, the size of the index, and pruning complex regions of the graph.

Mapping reads to a graph was first investigated by Schneeberger et~al.~\cite{Schneeberger2009}. Sirén et~al.~\cite{Siren2014} developed \emph{GCSA}, which generalized the FM\nobreakdash-index \cite{Ferragina2005a} from a text index to a path index. The original GCSA could only index directed acyclic graphs. It depended on pruning the complex regions, as there were no limits on query length. Kim et~al.~\cite{Kim2015-2016} combined GCSA with the HISAT read aligner \cite{Kim2015}. The result was the first practical graph-based read aligner, though it uses the graph for more accurate mapping to a linear reference.

The \emph{succinct de~Bruijn graphs} of Bowe et.~al.~\cite{Bowe2012} is generalization of the FM\nobreakdash-index as GCSA for representing \emph{de~Bruijn graphs} space-efficiently. Rødland \cite{Roedland2013} proposed another similar generalization. Succinct de~Bruijn graphs can represent \orderk{k} de~Bruijn graphs with different values of $k$ in the same structure \cite{Boucher2014}, but they still need to store the graph for the largest value of $k$ explicitly. Compacted \cite{Cazaux2014} and compressed de~Bruijn graphs \cite{Marcus2014} achieve similar space usage by representing unary paths in the graph as single nodes. A prototype read aligner based on compacted de~Bruijn graphs has been proposed \cite{Limasset2015}.

Other approaches to indexing variation graphs include BWBBLE \cite{Huang2013}, which indexes the graph as a collection of sequences. Simple variants are encoded with the powerset alphabet, while more complex variants are represented as separate sequences, with a sufficient amount of context around the variant site. Query times are much higher than in ordinary FM\nobreakdash-indexes due to the powerset alphabet. The hypertext index \cite{Thachuk2013} assumes that the nodes of the graph are labeled with strings. The strings are indexed using an ordinary FM-index. Partial matches in the strings are combined into full matches with range queries in the edge matrix. While matches crossing one edge are easy to find, the approach becomes impractical with matches crossing multiple edges. There are also several structures (e.g.~\cite{Huang2010,Wandelt2013,Danek2014,Na2015}), which use a graph as a space-efficient way of indexing similar sequences.

@@

A \emph{string} $S[0, n-1] = s_{0} \dotsm s_{n-1}$ of length $\abs{S} = n$ is a sequence of \emph{characters} over an \emph{alphabet} $\Sigma = \set{0, \dotsc, \sigma - 1}$. We often consider \emph{text} strings $T[0, n-1]$ terminated by an \emph{endmarker} $T[n-1] = \$ = 0$ not found anywhere else in the text. \emph{Binary} sequences are strings over the alphabet $\set{0, 1}$. A \emph{substring} of string $S$ is a sequence of the form $S[i, j] = s_{i} \dotsm s_{j}$. We call substrings of the type $S[0, j]$ and $S[i, n-1]$ \emph{prefixes} and \emph{suffixes}, respectively, and substrings of length $k$ as \kmer{k}s. Substring $S[i, j]$ is a \emph{proper} substring of string $S$, if $S \ne S[i, j]$. We say that string $S'$ is a substring of string collection $\mathcal{S}$, if it is a substring of a string $S \in \mathcal{S}$.

@@

Given string $S[0, n-1]$, we define $S.\rank(i, c)$ as the number of occurrences of character $c$ in the prefix $S[0, i-1]$. We also define $S.\select(i, c) = \arg \max_{j \le n} S.\rank(j, c) < i$ as the position of the occurrence of character $c$ with rank $i > 0$.\footnote{These definitions are used in the SDSL library \cite{Gog2014b}. We assume for convenience that $S.\select(0, c) = -1$.} A \emph{bitvector} is a binary sequence supporting efficient $\rank$/$\select$ queries. \emph{Wavelet trees} \cite{Grossi2003} are space-efficient data structures that use bitvectors to support $\rank$/$\select$ queries on arbitrary strings.

@@

The \emph{suffix tree} \cite{Weiner1973} is the most fundamental full-text index supporting substring queries. It is formed by taking the suffixes of the text, storing them in a trie, and compacting unary paths in the trie into single edges. If $v$ is a node of a suffix tree, we write $\ell(v)$ to denote the label of the path from the root to node $v$. Although fast and versatile, suffix trees are impractical with large texts, as they require much more space than the text itself.

\emph{Suffix arrays} \cite{Manber1993} were introduced as a space-efficient alternative to the suffix tree. The suffix array of text $T[0, n-1]$ is an array of pointers $\SA[0, n-1]$ to the suffixes of the text in \emph{lexicographic order}. We can find the occurrences of \emph{pattern} string $X$ in the text in $\Oh{\abs{X} \log n}$ time by using \emph{binary search} in the suffix array. The suffix array requires $n \log n$ bits of space in addition to the $n \log \sigma$ bits used by the text, while its functionality is more limited than that of the suffix tree. See Figure~\ref{figure:indexes} for an example of the suffix array and related structures.

@@

The \emph{Burrows-Wheeler transform (BWT)} \cite{Burrows1994} is a permutation of the text. Given text $T[0, n-1]$, its BWT is a string $\BWT[0, n-1]$, where $\BWT[i] = T[(\SA[i]-1) \bmod n]$. Given the \emph{lexicographic rank} $i$ of suffix $T[\SA[i], n-1]$, we can use \emph{\LFmapping} on the BWT to find the lexicographic rank of the previous suffix $T[(\SA[i]-1) \bmod n, n-1]$. Let
$$
\LF(i) = \Carray[\BWT[i]] + \BWT.\rank(i, \BWT[i]),
$$
where $\Carray[c]$ is the number of occurrences of characters $c' < c$ in the BWT. Then $\SA[\LF(i)] = (\SA[i]-1) \bmod n$. We can also generalize the definition for any character $c \in \Sigma$:
$$
\LF(i, c) = \Carray[c] + \BWT.\rank(i, c).
$$
Let $X$ be a string. If there are $i$ suffixes $S'$ of text $T$ such that $S' < X$ in lexicographic order, then there are $\LF(i, c)$ suffixes smaller than string $cX$ in lexicographic order.

We can use the BWT as a space-efficient text index. The \emph{\FMindex} \cite{Ferragina2005a} combines a representation of the BWT supporting $\rank$/$\select$ queries, the $\Carray$ array, and a set of \emph{sampled pointers} from the suffix array. It finds the \emph{lexicographic range} of suffixes matching pattern $X$ (having $X$ as a prefix) with a process called \emph{backward searching}. If the lexicographic range matching suffix $X[i+1, \abs{X}-1]$ of the pattern is $\SA[sp, ep]$, the range matching suffix $X[i, \abs{X}-1]$ is $\SA[\LF(sp, X[i]), \LF(ep+1, X[i]) - 1]$. Finding the lexicographic range matching the entire pattern requires $\Oh{\abs{X}}$ $\rank$ queries.

We use the sampled suffix array pointers for finding the text positions containing the occurrences. If $\SA[i]$ is not sampled, we start iterating $\LF(i)$, until we find a sampled pointer. If we find a sample at $\SA[\LF^{k}(i)]$, we know that
$$
\SA[i] = (\SA[\LF^{k}(i)] + k) \bmod n.
$$
If we have sampled one out of $d$ suffix array pointers at regular intervals, finding each occurrence takes $\Oh{d}$ rank queries. If we also sample one out of $d'$ \emph{inverse suffix array} pointers\footnote{The suffix array is a permutation of $\set{0, \dotsc, n-1}$, and the inverse suffix array is the inverse permutation.}, we can \emph{extract} an arbitrary substring $X$ of the text using $\Oh{\abs{X}+d'}$ rank queries.

@@

The graphs we use are \emph{labeled} with alphabet $\Sigma$: each node $v \in V$ has a \emph{label} $G.\glabel(v) \in \Sigma$. A \emph{path} in a graph is a sequence of nodes $P = v_{0} \dotsm v_{\abs{P}-1}$ such that $(v_{i}, v_{i+1}) \in E$ for $0 \le i < \abs{P}-1$. We say that $v_{0}$ is the \emph{start} node and $v_{\abs{P}-1}$ is the \emph{end} node of the path. The label of a path is the concatenation of node labels $G.\glabel(P) = G.\glabel(v_{0}) \dotsm G.\glabel(v_{\abs{P}-1})$.

@@

We assume that all graphs have two special nodes: the \emph{source} node $s$ and the \emph{sink} node $t$. To distinguish them from the other nodes, we label them with characters $G.\glabel(s) = \#$ and $G.\glabel(t) = \$$. These characters are only used to label technical nodes similar to nodes $s$ and $t$. We add an edge $(s, v)$ for all nodes $v \in V \setminus \set{s}$ that have no other incoming edges, and an edge $(v, t)$ for all nodes $v \in V \setminus \set{t}$ with no other outgoing edges. We also add edge $(t, s)$ to guarantee that $G.\gindegree(v) \ge 1$ and $G.\goutdegree(v) \ge 1$ for all nodes $v \in V$. However, this edge is not considered a real edge, and no path can cross it.

@@

We work with de~Bruijn graphs and their generalizations. For that purpose, we define collections of (finite or infinite) sequences suitable for constructing \orderk{k} de~Bruijn graphs.

@@

De~Bruijn graphs have several properties that make them useful for indexing purposes. Node keys prefix-match the labels of all paths starting from the node. This makes it possible to sort the nodes unambiguously by path labels. Every substring of the \kcollection{k} is the label of a path in the de~Bruijn graph, and every path label of length at most $k+1$ is a substring of the collection. In Section~\ref{sect:path-indexes}, we develop an index structure based on a generalization of de~Bruijn graphs.

@@

The \emph{generalized compressed suffix array} (GCSA) \cite{Siren2014} is a further generalization of the XBW transform for a class of graphs that includes \emph{directed acyclic graphs} and de Bruijn graphs. Before a graph can be indexed, we have to transform it into an equivalent graph, where the nodes can be unambiguously sorted by the labels of the right-maximal paths starting from them. The transformation increases the size of the graph exponentially in the worst case. After sorting the nodes of the transformed graph in lexicographic order, we encode them using three sequences. $\BWT$ contains the labels of the predecessors, while bitvectors $\bvIN$ and $\bvOUT$ encode the indegrees and outdegrees of each node in the same way as bitvector $B$ of the XBW transform. \LFmapping{} uses $\select$ queries on bitvector $\bvIN$ to map nodes to BWT ranges, ordinary \LFmapping{} with $\BWT$ to map incoming edges to the corresponding outgoing edges, and $\rank$ queries on bitvector $\bvOUT$ to map the outgoing edges to the predecessor nodes.

@@

The design of a path index is a trade-off between maximum query length and the size of the index. In the worst case, a graph may have up to $\sigma^{k}$ distinct labels of paths of length $k$ (ignoring the special treatment of characters $\#$ and $\$$), and cyclic graphs have paths of unlimited length. GCSA \cite{Siren2014} tried to avoid this by considering only a relatively simple class of graphs, using \emph{pruning heuristics} to simplify the graphs when necessary. The price of this was the possibility of \emph{false negatives}: path labels that exist in the graph but not in the index.

@@

Another representation trades query performance for the ability to search for short patterns. We sort the key-value pairs by their keys and store them in array $A$. If $X \in \Sigma^{\ast}$ is a pattern with $\abs{X} \le k$, we find the range $A[sp, ep]$ of pairs $(X_{i}, V_{i})$, where $X$ is a prefix of the key $X_{i}$, with binary search. We then list the union $\bigcup_{i=sp}^{ep} V_{i}$ as the set of matches.

We can use de~Bruijn graphs and \kmer{k} indexes to simulate each other. For that purpose, we define the de~Bruijn graph of graph $G = (V, E)$ by using the collection $\mathcal{S}$ of the labels of the maximal paths in the graph. If sequence $S \in \mathcal{S}$ is the label of path $P = (v_{i})_{i \in Z}$ in the graph, we set $\mathcal{S}.\gnode(S, i) = v_{i}$ for all positions $i \in Z$. We transform $\mathcal{S}$ into a \kcollection{k} by inserting characters $\#$ to the beginning of each non-left-infinite sequence when necessary, and characters $\$$ to the end of each non-right-infinite sequence. If $S[i]$ is a $\#$ we inserted, we set $\mathcal{S}.\gnode(S, i) = s:j$, where $s \in V$ is the start node and $j$ is the distance to the nearest non-inserted $\#$ in $S$. Similarly, if $S[i]$ is an inserted $\$$, we set $\mathcal{S}.\gnode(S, i) = (t, j)$, where $t$ is the sink node and $j$ is the distance to the nearest non-inserted $\$$.

\begin{definition}[de~Bruijn graph of a graph]\label{def:dbg}
Let $G$ be a labeled graph, and let $\mathcal{S}$ be the \kcollection{k} of maximal path labels in $G$. The \orderk{k} de~Bruijn graph of $\mathcal{S}$ is the \orderk{k} de~Bruijn graph of graph $G$.
\end{definition}

A \kmer{(k+1)} index of graph $G = (V, E)$ can simulate the \orderk{k} de~Bruijn graph $G' = (V', E')$ of $G$. We represent the nodes of the de~Bruijn graph with the adjacent edges. If we are interested in node $v \in V'$ with $G'.\gkey(v) = X$, we search for incoming edges $cX$ and outgoing edges $Xc$ for all $c \in \Sigma$. Determining the existence of a node requires $2 \sigma$ queries in the \kmer{k+1} index, and we learn the presence of adjacent nodes and edges in the process.

We can also use the \orderk{k} de~Bruijn graph $G' = (V', E')$ of graph $G = (V, E)$ as a \kmer{k} index of $G$. Let $\mathcal{S}$ be the \kcollection{k} in Definition~\ref{def:dbg}. We attach a set of nodes of graph $G$ to each node $v' \in V'$ as a \emph{value}: $G'.\gvalue(v) = \set{ \mathcal{S}.\gnode(S, i) \mid S \in \mathcal{S}, S[i, i+k-1] = G'.\gkey(v)}.$ Apart from some technicalities near the source/sink nodes, set $G'.\gvalue(v)$ is the set of start nodes of the paths $P$ matching pattern $G'.\gkey(v)$ in graph $G$. There will be no false negatives, regardless of pattern length. There may be \emph{false positives} (path labels that exist in the index but not in the input graph) with patterns longer than $k$, but we can avoid them by \emph{verifying} the results of such queries in the input graph. See Figure~\ref{figure:graph-dbg} for an example.

@@

@@

Path graphs generalize de~Bruijn graphs by allowing node keys of different lengths.

\begin{definition}[Path graph]
Let $\mathcal{S}$ be a \kcollection{k} of the labels of maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of substrings of length $k$ or less from $\mathcal{S}$. Assume that each right-maximal subtring $S \in \mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$, and that $\#^{k}, \$^{k} \in \mathcal{K}$. The \orderk{k} \emph{path graph} of graph $G$ with \emph{key set} $\mathcal{K}$ is a graph $G' = (V', E')$, where
\begin{itemize}
\item each node $v_{K} \in V'$ represent a distinct key $K \in \mathcal{K}$, with $G'.\glabel(v_{K}) = K[0]$;
\item each node has a key $G'.\gkey(v_{K}) = K$ and a value $G'.\gvalue(v_{K}) = V_{K}$, where $V_{K}$ is the set of nodes $\mathcal{S}.\gnode(S, i) \in V$ for $S \in \mathcal{S}$ and positions $i$ such that $S[i, i+\abs{K}-1] = K$; and
\item each edge $(v_{K}, v_{K'}) \in E'$ represents the occurrence of substring $K[0] K'$ in $\mathcal{S}$ such that strings $K$ and $K[0] K'$ prefix-match.
\end{itemize}
We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
\end{definition}
